A Comparison of Machine Learning Models for Limnological Time Series Forecasting

### 2024-02-27 manuscript/manuscript.Rmd
:pen:

The ability to accurately forecast water temperature, dissolved oxygen and chlorophyll concentration in freshwater systems is important to natural resource management. Since water temperature, dissolved oxygen and chlorophyll concentration effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive forecasts of these variables could allow managers to prepare for scenarios with far-reaching negative impacts. Limnologists have historically used empirical and process-based models to forecast water quality metrics, but in recent years, researchers have shown that machine learning can outperform these traditionally used methods \cite{Hanson, others}. A shortcoming of the limnological studies that support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of water systems across a continental scale. This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by examining how these models perform at 34 different sites across North America at different times of the year.

The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{Quinn}. A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{Dietze, Quinn}. For instance, in a model comparison that examined the forecasts for the phenology data from the Ecological Forecasting Challenge, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model. Thus, a primary consideration in this paper is the comparing the performance of the machine learning models to the climatology model.

There are many neural network architectures that have not been explored for ecological time series forecasting. Past applications of machine learning to limnological time series forecasting have typically used the long-short-term-memory (LSTM) neural network which was first introduced in 1997 \cite{lstm}. Over the nearly 30 years since LSTM's were introduced, there has been much development in the application of machine learning to time series forecasting, but recent studies examining machine learning for ecological time series forecasting have not explored more contemporary neural network architectures. This paper will compare 8 machine learning algorithms including LSTM's to ML methods that have been recently developed for time series forecasting like Temporal Fusion Transformer's and Temporal Convolutional Networks \cite{tft, tcn}.

@@

We used ML models to generate water temperature, dissolved oxygen and chlorophyll-A concentration forecasts for 34 freshwater sites across North America. These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes designated by NEON. We trained 8 different ML models for every site and target variable available under the Aquatics theme of the NEON Ecological Forecasting Challenge. For the models that accept past covariates (**list these models**), we inputted as covariates the other target variables recorded at that site as well as air temperature. For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate. The collected time series were split into a training and a validation set, whereby the models were trained on time series from 2018 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.

We compared the ML models to a climatology and naive persistence null model. The climatology model generates forecasts by finding the daily mean and standard deviation of the target time series, and then drawing samples from a Gaussian Distribution with these parameters. The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast window. We evaluate the continuous rank probability scores for the climatology model, and the absolute error for the Naive Persistence model.

@@

We provide fully reproducible code for fitting, scoring and visualizing the forecasts. All the machine learning forecasts were created using the `darts` python library. `darts` is similar to other libraries `scikit-learn` in Python or `tidymodels` in R that allow the user to use complex time series forecasting models without having to implement them. Otherwise, we rely on the `CRPS` library in python to efficiently compute the continuous ranked probability scores (CRPS). When comparing the probabilistic forecasts, we express the CRPS score in error-orientation -- i.e., a lower CRPS indicates a smaller margin of error.

@@

Plots. Talk about plots.

>>>>

The ability to accurately forecast water temperature, dissolved oxygen and chlorophyll concentration in freshwater systems is important to natural resource management. Since water temperature, dissolved oxygen and chlorophyll concentration effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive forecasts of these variables could allow managers to prepare for scenarios with far-reaching negative impacts. Limnologists have historically used empirical and process-based models to forecast water quality metrics, but in recent years, researchers have shown that machine learning can outperform these traditionally used methods \cite{Hanson, others}. A shortcoming of the limnological studies that support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of water systems on a broad geographic scale. This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by examining how these models perform at 34 different sites across North America at different times of the year.

The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{Quinn}. A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{Dietze, Quinn}. For instance, in a model comparison that examined the forecasts for phenology time series, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model. Thus, a primary consideration in this paper is comparing the performance of the machine learning models against the climatology model.

There are many neural network architectures that have not been explored for ecological time series forecasting. Past applications of machine learning to limnological time series forecasting have typically used the long-short-term-memory (LSTM) neural network which was first introduced in 1997 \cite{lstm}. Over the nearly 30 years since LSTM's were introduced, there has been much development in the subfield of machine learning for time series forecasting, but recent studies examining machine learning for ecological time series forecasting have not explored more contemporary neural network architectures. This paper will compare 8 machine learning algorithms including LSTM's to ML methods that have been recently developed for time series forecasting like Temporal Fusion Transformer's and Temporal Convolutional Networks \cite{tft, tcn}.

@@

We employed ML models to generate water temperature, dissolved oxygen and chlorophyll-A concentration forecasts for 34 freshwater sites across North America. These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON. 8 different ML models were trained for every site and target variable available in NEON Ecological Forecasting Challenge Aquatics theme. For the models that accept past covariates (**list these models**), we used the other target variables recorded at that site as well as air temperature as covariates. For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate. The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2018 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.

We compared the ML models to a climatology and naive persistence null model. The climatology model generates forecasts by finding the daily mean and standard deviation of the target time series, and then drawing samples from a Gaussian Distribution with these parameters. The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast window.

@@

We provide fully reproducible code for fitting, scoring and visualizing the forecasts. All the machine learning forecasts were generated using the `darts` python library. `darts` is similar to other libraries `scikit-learn` in Python or `tidymodels` in R that allow users to easily employ time series forecasting models without having to implement them. For model comparison, we rely on the `CRPS` library in Python to efficiently compute the continuous ranked probability scores (CRPS). When comparing the probabilistic forecasts, we express the CRPS score in error-orientation -- i.e., a lower CRPS indicates a smaller margin of error.

@@
